Data Imputation as a Query Optimization

### 2016-12-11 paper/algorithm.tex
paper fix: extract algorithm begin/end markers to be able to use with [H] placement in slides
paper: update imputation algorithm section

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric which combines the cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans $Q$ using a cost model $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$. $\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates. $\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance. $\alpha = 1.0$ means that the query should be as fast as possible, $\alpha=0.0$ means that the query should be as accurate as possible.

In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan refers to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in any modern RDBMS.\@ Our runtime cost model is perhaps less sophisticated than most, as it is identical to the cost model in SimpleDB.

$\text{Loss}(Q)$ is computed as follows.
For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered. First, all filters are pushed to the leaves of the query tree, immediately after the scans. Joins are performed after filtering, and only left-deep plans are considered. Any group-by/aggregate will be performed after the final join. Finally, projections are placed at the root of the query tree. The space of query plans is similar to that considered by System R\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

@@

\subsection{Imputation Strategies}
The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT)\cite{burgette2010multiple}. Note that plugging in a new
imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.

\begin{algorithm}
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of $T$ attributes that need to be imputed, and $C$ is a set of $T$ attributes that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCERT}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
			\State $imp \gets \Call{TrainRT}{\pi_C(T'), \pi_d{T'}}$
			\State $T' \gets (\pi_C(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for chained imputation using regression trees}
  \label{algo:imputation-strategy}
\end{algorithm}

\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}

>>>>

At a high level, our goal is to select a query plan that minimizes a metric which combines the cost of the query and the quality of the results.
Precisely, we will find a query plan $Q$ that minimizes $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$. $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance. $\alpha = 1.0$ means that the query should be as fast as possible, $\alpha=0.0$ means that the query should be as accurate as possible.

\subsection{Search space}
To reduce the size of the query plan search space, only plans that fit the following template are considered. First, all filters are pushed to the leaves of the query tree, immediately after the scans. Joins are performed after filtering, and only left-deep plans are considered. Any group-by/aggregate will be performed after the final join. Finally, projections are placed at the root of the query tree. The space of query plans is similar to that considered by Selinger \todo{Selinger cite}, with the addition of imputation operators appearing before/after traditional operators.

@@

\subsection{Histogram computation}
In order to correctly estimate the imputation cost and computation time parameters in our cost model,  the system must have updated cardinality estimates
at each point in a query plan. These cardinality estimates can be impacted not just be filtering or joining, as in the traditional relational calculus, but also by
the imputation operators. For example,  dropping versus imputing tuples has a clearly different effect on cardinality. For simplicity of design,
each of the logical nodes in a query plan keeps a reference to a set of histograms. The optimizer creates a copy-on-write of the histograms and modifies them
as necessary to account for new operations in the plan. In all cases, modifications maintain as an invariant that the distribution of data in the new histogram
matches the distribution (but not necessarily the cardinality) of the original statistics. Algorithm~\ref{algo:histogram-transformation} provides the details
of our implementation.

@@

\subsection{Imputation Strategies}

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data} and others) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2011multiple}.

These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
perform statistical analysis of the complete data. To this end, the technique of
\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
data are generated and estimators are computed by averaging over the multiple datasets. In
ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
note that in further work, multiple imputation could be used within ImputeDB to generate
lower-variance estimates of some aggregates, like averages.

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (the imputed values do not change across
epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

It is worth noting that given the nature of the imputation strategy used here, imputation
operators become blocking in our system. An extension of ImputeDB could consider regression
algorithms that learn in an online nature; this would reduce the cost of the imputation at
the possible expense of quality.

\begin{algorithm}
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of attributes of $T$ that need to be imputed, and
      $C$ is a set of attributes of $T$ that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCE-CART}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
            \State $imp \gets \Call{TrainRT}{\pi_{C \cup D \backslash \{d\}}(T'), \pi_d{T'}}$
            \State $T' \gets (\pi_{C \cup D \backslash \{d\}}(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for chained imputation using regression trees}
  \label{algo:imputation-strategy}
\end{algorithm}

\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
feel free to modify otherwise.}

### 2016-12-11 paper/conclusion.tex
paper: update imputation algorithm section

\subsection{Future Work}
\ProjectName{} opens up multiple avenues for further work. For instance, we could extend our minimal/maximal imputation
operators to consider global information, such as the specific columns needed in operators higher up in the query plan.
Our optimizer uses the same machine learning algorithm in all instances of imputation operators. Integrating multiple
possible algorithms could allow for further fine-grained imputation. Finally, the underlying database for \ProjectName{}
is far from a full-featured, production database, so a natural step is to integrate imputation planning in a widely-adopted
production-quality database. This will likely expose further opportunities for improvement.

>>>>

\subsection{Future Work}
\ProjectName{} opens up multiple avenues for further work. For instance, we
could extend our minimal/maximal imputation operators to consider global
information, such as the specific columns needed in operators higher up in the
query plan.  Our optimizer uses the same machine learning algorithm in all
instances of imputation operators. Integrating multiple possible algorithms
could allow for further fine-grained imputation. For example, algorithms that
learn in an online manner could increase the efficiency of the system.
Furthermore, a multiple imputation strategy could be followed for queries
involving certain aggregates. Finally, the underlying database for
\ProjectName{} is far from a full-featured, production database, so a natural
step is to integrate imputation planning in a widely-adopted production-quality
database. This will likely expose further opportunities for improvement.
